\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-07 paper/algorithm.tex
removed references to SimpleDB, clarified some stuff, and added TODOs

The space of query plans is similar to that considered by System R~\cite{blasgen1981system}, with the addition of imputation operators appearing before/after traditional operators.

@@

In order to correctly estimate $\textsc{Time}(Q)$ and $\textsc{Loss}(Q)$, the system must have cardinality estimates for each sub-query in each query plan.
These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.
For simplicity, each of the logical nodes in a query plan points to a set of histograms.
When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for new operation in the plan.
Algorithm~\ref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans.

Our runtime cost model is the cost model used in SimpleDB, though modern systems
have more sophisticated approaches.

@@

To compute $\text{Time}(Q)$, we retain the heuristics used by SimpleDB for the time
complexity of the typical relational operators, and add new formulations for \textit{Drop}
and \textit{Impute}. \textit{Drop} is a special case of a sequential scan, and its time
complexity can be expressed as a function of the number of heap pages read and the IO cost
per page.

@@

\subsection{Complexity}
Our optimization algorithm builds off the approach taken by System R \cite{blasgen1981system}, therefore our algorithm still operates in exponential time. Indeed, 
note that if we remove our restriction on types of imputation($\delta_{min}, \mu_{min}, \mu_{max}$), and allow any arbitrary subset of attributes to be imputed,
then every single operator in the query plan has a number of imputations exponential in the number of dirty columns. Our restriction, instead, increases the number
of plans (in the worst case) at each operator by a factor of 3. Of course, this implies that in the worst case (where all dirty sets tracked are distinct throughout the query plan),
we explore a number of plans that further scales the number of plans in the original algorithm by an exponential factor.

>>>>

The space of query plans is similar to that considered by System R~\cite{blasgen1981system}, with the addition of imputation operators appearing before/after traditional operators.
\todo{add a schematic of a standard query plan}

@@

In order to correctly estimate $\textsc{Time}(Q)$ and $\textsc{Loss}(Q)$, the system must have cardinality estimates for each sub-query in each query plan.
These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.
For simplicity, each of the logical nodes in a query plan points to a set of histograms.
When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for new operation in the plan.
Algorithm~\ref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans

@@

To compute $\text{Time}(Q)$, we retain a simple set of heuristics used by a standard 
database for the time. Scan costs are estimated based on the number of tuples, 
page size, and an unit-less IO cost per page. Join costs are estimated as a function
of the two relations joined, any blocking operations, and repeated fetching of tuples (if required).
Filtering, projecting, and aggregating are all done in memory and have low computational overhead
so we assume negligible time costs for those operations. We extend these heuristics to 
account for the new operators: \textit{Drop} and \textit{Impute}. \textit{Drop} is a special case of a sequential scan, and its time
complexity can be expressed as a function of the number of heap pages read and the IO cost
per page.

@@

\subsection{Complexity}
\todo{this section needs to be cleaned up/clarified}
Our optimization algorithm builds off the approach taken by System R \cite{blasgen1981system}, therefore our algorithm still operates in exponential time. Indeed, 
note that if we remove our restriction on types of imputation($\delta_{min}, \mu_{min}, \mu_{max}$), and allow any arbitrary subset of attributes to be imputed,
then every single operator in the query plan has a number of imputations exponential in the number of dirty columns. Our restriction, instead, increases the number
of plans (in the worst case) at each operator by a factor of 3. Of course, this implies that in the worst case (where all dirty sets tracked are distinct throughout the query plan),
we explore a number of plans that further scales the number of plans in the original algorithm by an exponential factor.

### 2017-02-07 paper/examples.tex
paper: formatting
removed references to SimpleDB, clarified some stuff, and added TODOs

Consider the case of an analyst looking to explore the sources and causes of
polling error in a presidential election. One hypothesis \cite{zukin2015s}
is that the rapid switch from landlines to cellphones increases the
difficulty of contacting potential poll respondents. The analyst may
begin her analysis by comparing the polling error in a state with the
distribution of households with landlines, for the set of states for which
polling error was most pronounced. In order to probe deeply, the analyst
accesses a household-level survey that includes data on
whether the household has a landline. One such survey --- which will be
discussed further (Sec~\ref{subsec:datasets}) --- is the American Community Survey (ACS),
conducted by the US Census Bureau.

Surveys are rife with missing values due to non-response and other issues, and
the ACS is no exception. The Census Bureau and other similar organization attempt to present
data in as unadulterated a fashion as possible. The ACS dataset is not amenable to one-time
imputation by either the Census Bureau, for this reason, or the analyst's organization,
which likely uses the same database for a variety of unrelated uses (asking for different
imputation strategies). Thus, the analyst must deal with missing values in one way or
another.

In this case, a non-negligible fraction of respondents omitted information on whether they
have a landline. The analyst hypothesizes that older houses are more likely to have
landlines and that households that are relatively new are less likely to have landlines. 
The analyst could take advantage of these presumed correlations to impute the missing values
on the base table in one expensive operation. On the other hand, the query could be run
directly and rows with missing values may be dropped entirely. With ImputeDB, the analyst
could impute the relevant subset of data on-the-fly, achieving better performance without
losing the information in the dropped examples.

The analyst submits a query (\todo{ref to listing}) to ImputeDB which finds an optimized query
plan\footnote{The analyst specifies a relatively high value of $\alpha$, the quality
emphasis parameter, in order to receive a plan that does an expensive imputation step. See
Section~\ref{sec:cost-model} for more details.}. The resulting query plan
(Figure~\ref{fig:query-plan})
imputes missing values for \verb!acs.TEL! after states with high polling error have been
selected. This decision reduces the set of tuples that must be input into the imputation
algorithm to just those that are relevant to the query. Later, we show the algorithms that
the ImputeDB query optimizer uses in order to detect these opportunities and control the
tradeoffs between efficiency and imputation quality.

\begin{verbatim}
SELECT polling.ST, AVG(acs.TEL)
FROM polling, acs
WHERE polling.ST = acs.ST
  AND polling.ERROR > 50         -- 5 percentage points
GROUP BY polling.ST;
\end{verbatim}

\begin{figure}[!ht]
    \centering
    \includegraphics[width=0.6\textwidth]{figures/example.png}
    \caption{Query plan generated by ImputeDB prioritizing imputation quality.}
    \label{fig:query-plan}
\end{figure}

>>>>

\todo{Replace example with a relevant example and plan to the new data used for experiments}
%Consider the case of an analyst looking to explore the sources and causes of
%polling error in a presidential election. One hypothesis \cite{zukin2015s}
%is that the rapid switch from landlines to cellphones increases the
%difficulty of contacting potential poll respondents. The analyst may
%begin her analysis by comparing the polling error in a state with the
%distribution of households with landlines, for the set of states for which
%polling error was most pronounced. In order to probe deeply, the analyst
%accesses a household-level survey that includes data on
%whether the household has a landline. One such survey --- which will be
%discussed further (Sec~\ref{subsec:datasets}) --- is the American Community Survey (ACS),
%conducted by the US Census Bureau.
%
%Surveys are rife with missing values due to non-response and other issues, and
%the ACS is no exception. The Census Bureau and other similar organization attempt to present
%data in as unadulterated a fashion as possible. The ACS dataset is not amenable to one-time
%imputation by either the Census Bureau, for this reason, or the analyst's organization,
%which likely uses the same database for a variety of unrelated uses (asking for different
%imputation strategies). Thus, the analyst must deal with missing values in one way or
%another.
%
%In this case, a non-negligible fraction of respondents omitted information on whether they
%have a landline. The analyst hypothesizes that older houses are more likely to have
%landlines and that households that are relatively new are less likely to have landlines. 
%The analyst could take advantage of these presumed correlations to impute the missing values
%on the base table in one expensive operation. On the other hand, the query could be run
%directly and rows with missing values may be dropped entirely. With ImputeDB, the analyst
%could impute the relevant subset of data on-the-fly, achieving better performance without
%losing the information in the dropped examples.
%
%The analyst submits a query (Figure~\ref{fig:example-query}) to ImputeDB which finds an optimized query
%plan\footnote{The analyst specifies a relatively high value of $\alpha$, the quality
%emphasis parameter, in order to receive a plan that does an expensive imputation step. See
%Section~\ref{sec:cost-model} for more details.}. The resulting query plan
%(Figure~\ref{fig:query-plan})
%imputes missing values for \verb!acs.TEL! after states with high polling error have been
%selected. This decision reduces the set of tuples that must be input into the imputation
%algorithm to just those that are relevant to the query. Later, we show the algorithms that
%the ImputeDB query optimizer uses in order to detect these opportunities and control the
%tradeoffs between efficiency and imputation quality. Finally, note that the optimizer doesn't
%place any imputation operator for the \verb|acs.ST| column, as it uses a histogram to
%detect that there are no missing values there (as might be expected for a survey question that isn't
%impacted by non-responses).
%
%\begin{figure}
%\begin{lstlisting}[language=SQL]
%SELECT polling.ST, AVG(acs.TEL)
%FROM polling, acs
%WHERE polling.ST = acs.ST
%  AND polling.ERROR > 50         -- 5 percentage points
%GROUP BY polling.ST;
%\end{lstlisting}
%\caption{A typical analyst query on ACS data}
%\label{fig:example-query}
%\end{figure}
%
%\begin{figure}[!ht]
%    \centering
%    \includegraphics[width=0.6\columnwidth]{figures/example.png}
%    \caption{Query plan generated by ImputeDB prioritizing imputation quality}
%    \label{fig:query-plan}
%\end{figure}

### 2017-02-07 paper/main.tex
paper: make paper shorter
paper: conform to vldb style
paper: formatting
removed references to SimpleDB, clarified some stuff, and added TODOs

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage[activate={true,nocompatibility},final,kerning=true,spacing=true,factor=1100,stretch=10,shrink=10]{microtype}
\usepackage{amsfonts,amssymb,amsmath}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[style=ieee]{biblatex}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{hyperref} % load after other packages
\usepackage{graphicx} % includegraphics
\usepackage{listings} % code samples with syntax hl
\addbibresource{references.bib}

@@

\author{
  Jose Cambronero \\
  \and
  John K. Feser
  \and
  Micah Smith}

@@

\begin{abstract}
In order to study the placement of an imputation step, we create a logical imputation operator (along with respective physical instances) and incorporate its placement as part of the query plan optimization process in SimpleDB.  We introduce measures for information loss and runtime for imputation operations, which outline the main trade-offs in the imputation placement. We add these measures into our cost estimation, allowing us to intelligently place the data imputation step during query planning. We show the trade-offs between efficiency and accuracy for simple data imputation models. \end{abstract}

@@

\printbibliography

>>>>

\documentclass{vldb}
\usepackage{balance}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage[activate={true,nocompatibility},final,kerning=true,spacing=true,factor=1100,stretch=10,shrink=10]{microtype}
\usepackage[subtle]{savetrees}
\usepackage{amsfonts,amssymb,amsmath}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[style=numeric,firstinits=true]{biblatex}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{hyperref} % load after other packages
\usepackage{graphicx} % includegraphics
\usepackage{listings} % code samples with syntax hl
\addbibresource{references.bib}

\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}

@@

\author{
  Jos\'e Cambronero \\
  \texttt{jcamsan@mit.edu}
  \and
  John K. Feser \\
  \texttt{feser@csail.mit.edu}
  \and
  Micah Smith \\
  \texttt{micahs@mit.edu}}

@@

\begin{abstract}
In order to study the placement of an imputation step, we create a logical imputation operator (along with respective physical instances) and incorporate its placement as part of the query plan optimization process into a simple database prototype.  We introduce measures for information loss and runtime for imputation operations, which outline the main trade-offs in the imputation placement. We add these measures into our cost estimation, allowing us to intelligently place the data imputation step during query planning. We show the trade-offs between efficiency and accuracy for simple data imputation models. \end{abstract}

@@

\balance
\printbibliography
